---
title: Using the `GenomicInteractions` class
author: 
- name: Aaron Lun
  email: infinite.monkeys.with.keyboards@gmail.com
- name: Malcolm Perry
- name: Elizabeth Ing-Simmons
date: "Revised: 6 May 2019"
output:
  BiocStyle::html_document:
    toc_float: yes
package: GenomicInteractions    
vignette: >
  %\VignetteIndexEntry{1. The GenomicInteractions class}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignettePackage{UTF-8}
---    

```{r, echo=FALSE, results="hide", message=FALSE}
require(knitr)
opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
require(GenomicInteractions)
```

# Introduction

Techniques such as Hi-C and ChIA-PET have driven the study of genomic interactions, i.e., physical interactions between pairs of genomic regions.
The `r Biocpkg("GenomicInteractions")` package implements the `GenomicInteractions` class to represent and manipulate these interactions.
The aim is to provide package developers with stable class definitions that can be manipulated through a large set of methods.
It also provides users with a consistent and intuitive interface across different packages that use the same classes, making it easier to perform analyses with multiple packages.

# Construction 

The `GenomicInteractions` class stores any number of pairwise interactions between genomic regions.
The regions themselves are represented by a `GenomicRanges` object from the `r Biocpkg("GenomicRanges")` package.
For example, consider an `all.regions` object containing consecutive genomic intervals^[While any regions can be used here, consecutive intervals are just simpler to explain.]:

```{r}
all.regions <- GRanges("chrA", IRanges(0:9*10+1, 1:10*10))
```

Let there be pairwise interactions between elements of `all.regions`.
We will consider three pairwise interactions -- one between region #1 and #3, another between #5 and #2, and the last between #10 and #6.

```{r}
index.1 <- c(1,5,10)
index.2 <- c(3,2,6)
region.1 <- all.regions[index.1]
region.2 <- all.regions[index.2]
```

Construction of a `GInteractions` object is performed by supplying the interacting regions to the `GenomicInteractions()` constructor:

```{r}
gi <- GenomicInteractions(region.1, region.2)
gi
```

This generates a `GenomicInteractions` object of length 3 where each entry corresponds to a pairwise interaction.
Alternatively, the indices can be supplied directly, along with the coordinates of the regions they refer to:

```{r}
gi <- GenomicInteractions(index.1, index.2, all.regions)
gi
```

**Advanced use:**
The `GenomicInteractions` class is derived from the `IndexedRelations` class from the `r Biocpkg("IndexedRelations")` package.
This makes use of indexing to avoid storing redundant copies of the same interactors.
Specifically, in a `GenomicInteractions` instance, `GenomicRanges` for each anchor are not stored explicitly.
Rather, only the unique entries are stored for each anchor, representing the "universe" of all possible regions for that anchor.
Each interaction then stores the indices to point at the relevant intervals in the universe for that anchor.

In many applications, the same intervals are re-used for different interactions, e.g., common bins in Hi-C data, common peaks in ChIA-PET data. 
Storing indices rather than repeated `GRanges` entries allows developers to write efficient algorithms that avoid redundant operations.
It also reduces memory usage though this is usually not a major consideration:

```{r}
big.1 <- sample(all.regions, 1e6, replace=TRUE)
big.2 <- sample(all.regions, 1e6, replace=TRUE)
big.gi <- GenomicInteractions(big.1, big.2)
object.size(big.gi)
object.size(Pairs(big.1, big.2)) # For comparison
```

# Extracting information

The interacting regions are referred to as anchor regions because they "anchor" the ends of the interaction^[Think of them like the cups in a string telephone.].
These anchor regions are accessible with the `anchors` method:

```{r}
anchors(gi)
```

This returns a `Pairs` object containing the pairs of anchor regions, where each pair represents a single interaction in `gi`.
We can also obtain `GRanges` for the first or second anchor regions by themselves by specifying `type`:

```{r}
anchors(gi, type=1)
anchors(gi, type=2)
```

The `first()` and `second()` methods can also be used to the same effect.

**Advanced use:**
When writing functions that operate on `GenomicInteractions` instances, it is often more efficient to manipulate the indices directly.
Indices are extracted by specifying `id=TRUE` in `anchors()`:

```{r}
anchors(gi, id=TRUE)
```

The universe of regions to which those indices point are obtained with the `regions()` method:

```{r}
regs <- regions(gi, type=NULL)
regs[[1]]
regs[[2]]
```

Common operations can then be applied to the universe of non-redundant regions, and the relevant results retrieved with the anchor indices.
This is usually faster than applying those operations on repeated instances of the regions in `anchors(gi)`.

## Setters 

Modification of the anchors in an existing `GenomicInteractions` object is performed with the `anchors()<-` replacement method.
The code below re-specifies the three pairwise interactions as that between regions #1 and #5; between #2 and #6; and between #3 and #7.

```{r}
temp.gi <- gi
anchors(temp.gi, type=1) <- all.regions[1:3]
anchors(temp.gi, type=2) <- all.regions[5:7]
temp.gi
```

**Advanced use:**
Indices are directly modifiable by using `anchors()<-` with `id=TRUE`.

```{r}
temp.gi <- gi
anchors(temp.gi, type=1, id=TRUE) <- 1:3
temp.gi
```

The universe of regions for each anchor is modifiable directly using `regions()<-`.
One typical application would be to annotate regions with some metadata, 
e.g., GC content, surrounding genes, whether or not it is a promoter or enhancer:

```{r}
temp.gi <- gi
annotation <- rep(c("E", "P", "N"), length.out=length(all.regions))
regions(temp.gi, type=1)$anno <- annotation
```

This annotation will propagate to all anchor regions when they are retrieved:

```{r}
anchors(temp.gi, type=1)
```

# Subsetting and combining

A `GenomicInteractions` instance follows vector-like semantics where each interaction is an element.
Subsetting of a `GInteractions` object will return a new object containing only the specified interactions:

```{r}
gi[1:2]
```

Objects can also be concatenated using  `c`. 
This forms a new `GenomicInteractions` object that contains all of the interactions in the constituent objects.

```{r}
gi2 <- GenomicInteractions(all.regions[1:3], all.regions[5:7])
combined <- c(gi, gi2)
combined
```

**Advanced use:**
When combining constituent objects with different universes of regions, the output object will contain a "union of universes".
As a result, indices prior to combining may not refer to the same entries in the combined object.
Advanced users should make sure to obtain the new anchor indices after combining.

# Vector-like comparisons

Ordering of `GenomicInteractions` objects is performed based on the genomic position of the first anchor.
Any interactions with the same first anchor are ordered by the second anchor.

```{r}
order(combined)
sort(combined)
```

Duplicated interactions are identified as those that have identical pairs of anchor indices.
In the example below, all of the repeated entries in `doubled` are marked as duplicates.
The `unique` method returns a `GenomicInteractions` object where all duplicated entries are removed.

```{r}
doubled <- c(gi, gi)
duplicated(doubled)
unique(doubled)
```

Interactions are matched between `GenomicInteractions` objects using the `match()` method.
The example below will return the index of the first interaction in `combined` that matches each interaction in `gi`.

```{r}
match(gi, combined)
```

Finally, `GenomicInteractions` objects can be compared in a parallel manner.
This determines whether the $i$^th^ interaction in one object is equal to the $i$^th^ interaction in the other object.

```{r}
gi==gi2
```

# Inherited methods

The `GenomicInteractions` object inherits from the `IndexedRelations` class, which in turn inherits from `Vector` in the `r Biocpkg("S4Vectors")` package.
A `GenomicInteractions` instance has access to all of the methods written for its parent classes.
For example, general metadata can be dumped into a `GenomicInteractions` object using the `metadata` method:

```{r}
metadata(gi)$description <- "I am a GenomicInteractions object"
metadata(gi)
```

Interaction-specific metadata can be stored via the `mcols` replacement method or through the `$` wrapper^[Note that this differs from the region-specific metadata that was set using `regions()<-` above.].
One application might be to store interesting metrics relevant to each interaction, such as normalized contact frequencies:

```{r}
set.seed(1000)
norm.freq <- rnorm(length(gi)) # obviously, these are not real frequencies.
gi$norm.freq <- norm.freq
gi
```

**Advanced use:**
Developers may be interested in the suite of methods from the `r Biocpkg("IndexedRelations")` class.
These are explicitly designed for manipulating the indices and features for efficient operation of downstream algorithms.

```{r}
partnerNames(gi)

# Useful methods that work on GenomicInteractions objects.
X <- standardizeFeatureSets(gi, list(gi2))
X <- cleanFeatureSets(gi)
X <- dropUnusedFeatures(gi)
```

# Conclusion 

Details for any given method can be found through the standard documentation, 
e.g., `?anchors` to get the man page for the `anchors` method.
Further questions on `r Biocpkg("GenomicInteractions")` functionality should be asked on the [Bioconductor support site](https://support.bioconductor.org).

```{r}
sessionInfo()
```
