---
title: Performing overlaps with `GenomicInteractions` objects
author:
- name: Aaron Lun
  email: infinite.monkeys.with.keyboards@gmail.com
- name: Malcolm Perry
- name: Elizabeth Ing-Simmons
date: "Revised: 6 May 2019"
output:
  BiocStyle::html_document:
    toc_float: yes
package: GenomicInteractions
vignette: >
  %\VignetteIndexEntry{3. Overlapping interactions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignettePackage{UTF-8}
---

```{r, echo=FALSE, results="hide", message=FALSE}
require(knitr)
opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
library(BiocStyle)
require(GenomicInteractions)
```

# Introduction

The `r Biocpkg("GenomicInteractions")` package provides methods to overlap interactions based on their genomic coordinates. 
This is particularly useful for identifying interactions that are anchored at genomic intervals of interest.
or for identifying interactions that link two intervals of interest.
To demonstrate, we will create a mock `GenomicInteractions` object below:

```{r}
set.seed(9999)
all.regions <- GRanges(rep(c("chrA", "chrB"), c(10, 5)), 
    IRanges(c(0:9*10+1, 0:4*5+1), c(1:10*10, 1:5*5)))
index.1 <- sample(length(all.regions), 20, replace=TRUE)
index.2 <- sample(length(all.regions), 20, replace=TRUE)
gi <- GenomicInteractions(index.1, index.2, all.regions)
gi
```

# One-dimensional interactions

Say we want to identify all interactions with at least one anchor region lying within a region of interest (e.g., a known promoter or gene).
This is achieved with the `findOverlaps` method:
    
```{r}
of.interest <- GRanges("chrA", IRanges(30, 60))
olap <- findOverlaps(gi, of.interest)
olap
```

This returns a `Hits` object containing pairs of indices, where each pair represents an overlap between the interaction (`query`) with a genomic interval (`subject`).
Here, each reported interaction has at least one anchor region overlapping the interval specified in `of.interest`:

```{r}
anchors(gi[queryHits(olap)])
```

Longer `GRanges` can be specified if there are several regions of interest:

```{r}
of.interest2 <- GRanges(c("chrA", "chrB"),
    IRanges(c(30, 20), c(60, 30)))
findOverlaps(gi, of.interest2)
```

These overlaps are considered to be "one-dimensional" as they based on the linear genome.
By default, one-dimensional overlaps are considered for both anchor regions of each interaction.
The `use.region` argument can be set to specify which anchors in the `GenomicInteractions` object are to be overlapped.
For example, to only identify interactions where the first anchor overlaps `of.interest`:

```{r}
findOverlaps(gi, of.interest, use.region="first")
```

Standard arguments can be supplied to `findOverlaps` to modify its behaviour, e.g., `type`, `minoverlap`.
The `overlapsAny`, `countOverlaps` and `subsetByOverlaps` methods are also available and behave as expected.

# Two-dimensional interactions

A more complex situation involves identifying overlapping interactions in the two-dimensional interaction space.
Say we are interested in an interaction betweeen two regions, represented by an `GenomicInteractions` object named `paired.interest`.
We want to determine if any of our interactions in `gi` overlap with the existing interaction, e.g., to identify corresponding interactions between data sets.

This is achieved by passing both `GenomicInteractions` objects to `findOverlaps()`.
The function will only consider two interactions to overlap if each anchor region of the new interaction overlaps a corresponding anchor region of the existing interaction. 
To illustrate:

```{r}
paired.interest <- GenomicInteractions(of.interest, 
    GRanges("chrB", IRanges(10, 40)))
paired.interest
olap <- findOverlaps(gi, paired.interest)
olap
```

If we inspect the overlapping interactions, it is clear that interactions are only reported as overlapping if both of its anchors overlap the two anchors of `paired.interest`.
Specifically, the first `gi` anchor overlaps the first `paired.interest` anchor and the second `gi` anchor overlaps the second `paired.interest` anchor;
or the first `gi` anchor overlaps the second `paired.interest` anchor and the first `gi` anchor overlaps the second `paired.interest` anchor.

```{r}
gi[queryHits(olap)]
```

The `use.region` argument controls the regions that are used in the two-dimensional overlaps.
For example, setting `use.region="match"` ensures that only the first anchors are allowed to overlap each other, and only the second anchors are allowed to overlap each other:

```{r}
findOverlaps(gi, paired.interest, use.region="match")
```

Again, arguments can be supplied to `findOverlaps` to tune its behaviour.
The `overlapsAny`, `countOverlaps` and `subsetByOverlaps` methods are also available for these two-dimensional overlaps.

# 1.5-dimensional overlaps

The `findOverlaps()` method also supports one-dimensional overlaps between two `GenomicInteractions` objects.
(These are referred to as "1.5-dimensional" as they involve one-dimensional overlaps on two-dimensional objects.)
For example, to identify all overlaps of the first anchor regions between two `GenomicInteractions` objects, one would do:

```{r}
findOverlaps(gi, paired.interest, use.region="first-first")
```

This is exactly the same as, but more efficient than:

```{r}
findOverlaps(first(gi), first(paired.interest))
```

The efficiency is due to the fact that the 1.5-dimensional overlap avoids the need to expand out redundant entries of `gi` into a `GRanges` object.
The `findOverlaps()` function can exploit the underlying indexing structure of a `GenomicInteractions` object to avoid redundant work.

A variety of `use.region` options are available for 1.5-dimensional overlaps to control the anchor regions that are used.
All have the following structure:

- `use.region="first-first"` will identify all overlaps between the first anchors of the query object and the first anchors of the subject object.
- `use.region="first-second"` will identify all overlaps between the first anchors of the query object and the second anchors of the subject object.
- `use.region="first-any"` will identify all overlaps between the first anchors of the query object and either the first and second anchors of the subject object.
- `use.region="any-any"` will identify all overlaps between the any anchor of the query object and any anchor of the subject object.

And with a similar structure for `use.region="second-second"` and so on.

# Linking sets of regions 

A slightly different problem involves finding interactions that link any entries across two sets of regions.
For example, we might be interested in identifying interactions between a set of genes and a set of enhancers.
Using `findOverlaps()` to perform 2D overlaps would be tedious, as we would have to manually specify every possible gene-enhancer combination.
Instead, our aim can be achieved using the `linkOverlaps` function:

```{r}
all.genes <- GRanges("chrA", IRanges(0:9*10, 1:10*10))
all.enhancers <- GRanges("chrB", IRanges(0:9*10, 1:10*10))
out <- linkOverlaps(gi, all.genes, all.enhancers)
out
```

This returns a `DataFrame` object where each row species an interaction in `query`, the region it overlaps in `subject1` (i.e., the gene), 
    and the overlapping region in `subject2` (i.e., the enhancer).
If there are multiple overlaps to either set, all combinations of two overlapping regions are reported.
One can also identify interactions within a single set by doing:

```{r}
out <- linkOverlaps(gi, all.genes)
out
```

Here, both `subject1` and `subject2` refer to linked entries in `all.genes`.

# Wrapping up

Details for any given method can be found through the standard documentation, 
e.g., `?linkOverlaps` to get the man page for the `linkOverlaps()` method.
Further questions on `r Biocpkg("GenomicInteractions")` functionality should be asked on the [Bioconductor support site](https://support.bioconductor.org).

```{r}
sessionInfo()
```
